I would like to express my deepest gratitude to my research advisor, Dr. Bhaswar Ghosh, for his 
guidance and support. His mentorship has been invaluable to me and his expertise, patience, and 
encouragement have been instrumental in the successful completion of this thesis. \\

I am eternally indebted to my family for their unconditional love, encouragement, and sacrifices 
throughout my academic journey. In particular, I would like to thank my parents, grandparents and my sister, 
who have always been there for me to provide emotional support and wise counsel. \\

I would also like to acknowledge the exceptional support I have received from my closest friends. 
Anir, Arvindh, Ashu, Chandan, Dinesh, Jai, Jerrin, Jose, Lokesh, Manas, Rohan, Sanjai and Vijay in particular
have been a constant source of encouragement and laughter throughout my academic journey. 
Their unwavering friendship has made this journey all the more enjoyable. A special acknowledgement is to 
one of my first friends here, Harshit, for also being the best roommate one could ask for. \\

Last but certainly not least, I extend my heartfelt gratitude to IIIT Hyderabad for providing me with an 
exceptional platform for my academic and research endeavors. The institute has opened doors to incredible 
professional opportunities and, most importantly, allowed me to cross paths with so many remarkable people. 
As the saying goes, ``the hardest part of ending is starting again"; but I am eager to embark on the next 
chapter of my journey with the invaluable experiences and connections I gained here.
